---
title: "Workload Analysis"
output: html_notebook
---

## Workload in the age of COVID

Anecdotally, workload has been high for geographic branches within Global Affairs Canada as it wrestled with the ffects of a worldwide pandemic over most of 2020 and into 2021. Missions were particularly affected by repariation efforts and other elements of pandemic response, while HG staff provided support.

The 2020 Public Service Employee Survey provides us with an opportunity to look at how this has affected staff as of autumn 2020 - well into the second phase of the pandemic.

## Geograpic Branches vs the Rest of Global affairs Canada

Let us start by looking at how different geographic brancehs have done of workload-realted questions compared to toher branches.

### Loading our data

As usual, we start with loading our PSES data and a few packages.

```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(ggrepel)
library(tidymodels)
library(tidybayes)

source("load_pses_2020.R")

short_qs <- short_qs <- read_excel(file.path(data_dir,"pses_2020_short_questions.xlsx")) 

# Our department code (14 for Global Affairs Canada)
this_dept <-  14

```

### The right questions

Workload is a nebulous concept. There are many dimensions we coudl choose to look at. Let's start by seeing what the PSES can provide.

```{r}
pses_2020 %>% 
  filter(title_e %>% str_detect("workload|work hours|more work")) %>% 
  distinct(question, title_e)
```
This seems like a good start - 6 questions. We will exclude the COVID-specific questions for now.

```{r}
workload_qs <- c("Q17", "Q18e", "Q70c", "Q70f")
```


### Meaningful comparisons

The art of data visualization is based on making meaningful comparisons. How can we meaningfully compare the workload burden of geographic branches? We can start by doing both a within-group and out-of-group comparison. 

Let's identify our branches.

```{r}
pses_2020 %>% 
  filter(level1id == 14, level2id != 999, level3id == 0, subset == "ss7") %>% 
  distinct(descrip_e)
```



```{r, fig.width = 9, fig.height = 5}

workload_df <- pses_2020 %>% 
  filter(
    # Start by selecting only our level2 orgs - these are the branches at GAC
    level1id == 14, level2id != 999, level3id == 0, subset == "ss7",
    # Keep only workload questions
    question %in% workload_qs   
    ) %>%
  # Get our short questions. This helps with space!
  left_join(short_qs %>% select(question, q_short_e), by = "question") %>% 
  mutate(
    # Identify our geo branches (those that end with "GM")
    geo = case_when(
      descrip_e %>% str_detect("^\\wGM") ~ "Geo",
      level2id == 0 ~ "GAC",
      TRUE ~ "Not Geo"
    ),
    branch = case_when(
      level2id == 0 ~ "GAC",
      TRUE ~ descrip_e %>% str_sub(1,3)
    ) %>% fct_inorder() %>% fct_rev(),
    q_display = paste0(question, ". ", q_short_e) %>%
      str_wrap(25) %>%
      fct_reorder(positive)
  )

branch_clrs <- c(
   "black",
  viridis::viridis(4, begin = 0.2, end = 0.8)
) %>%
  set_names(
    workload_df %>%
      filter(geo != "Not Geo") %>% 
      distinct(branch) %>%
      pull()
  )

ggplot(mapping = aes(x = q_display, y = positive)) +
  geom_point(
    data = workload_df %>% filter(geo == "Not Geo"),
    colour = "grey80", alpha = 0.7
    ) +
  geom_line(
    data = workload_df %>% filter(geo == "Not Geo"),
    aes(group = branch),
    colour = "grey80", alpha = 0.5,
    orientation = "x", size = 1
  ) +
  geom_text_repel(
    data = workload_df %>% filter(question == "Q70f", geo != "Not Geo"),
    aes(label = branch, colour = branch, fontface = "bold"),
    direction = "y",
    nudge_x = 0.3
  ) +
  geom_point(
    data = workload_df %>% filter(geo != "Not Geo"),
    aes(colour = branch)
    ) +
  geom_line(
    data = workload_df %>% filter(geo != "Not Geo"),
    aes(group = branch, colour = branch),
    orientation = "x", alpha = 0.7, size = 1
  ) +
  scale_colour_manual(values = branch_clrs) +
  scale_y_continuous(limits = c(0,100), breaks = c(0,50,100)) +
  labs(
    title = "Geographic branches are less positive on workload questions",
    subtitle = "Percentage of positive responses to selected workload questions from PSES 2020.",
    caption = "Source: 2020 Public Service Employee Survey (https://open.canada.ca)",
    colour = NULL, x = NULL, y = NULL
    ) +
  theme_minimal_hgrid() +
  theme(
    legend.position = "none",
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank()
  )
  
ggsave(file.path(plot_dir,"PSES2020@GAC - Geographic Branches and Workload.png"))
```

We can also plot a regression.


```{r}

model_branch <- function(df = workload_df, q = NULL) {

#df = workload_df
#q = "Q70f"

#n_df <- df %>% 
#  filter(ee_dimension == ee_dim, seniority == level) %>%
#  count(ee_class)

log_mod <- logistic_reg() %>%
  set_engine("stan", cores = 24) %>% 
  set_mode("classification")

log_df <- df %>% 
  filter(question == q, geo != "GAC") %>%
  mutate(not_positive = 100 - positive) %>% 
  select(question, q_short_e, branch, geo, positive, not_positive, anscount) %>% 
  pivot_longer(
    cols = c(positive, not_positive), 
    names_to = "response", 
    values_to = "prop"
  ) %>% 
  mutate(count = round(prop/100 * anscount)) %>% 
  uncount(weights = count) %>% 
  mutate(
    geo = fct_drop(geo),
    branch = fct_drop(branch),
    response = as_factor(response)
  )

mult_fit <- log_mod %>%
  fit(
    response ~ geo + branch, 
    data = log_df
  ) 

saveRDS(mult_fit, file = file.path(output_dir, paste0("branch_workload_", "q",".rds")))
#broom.mixed::tidy(
#  mult_fit,
#  conf.int = TRUE,
#  conf.level = 0.8,
#  conf.method = "HPDinterval"
#)

branch_values <- log_df %>% 
  distinct(question, q_short_e, geo, branch)


branch_pred <- 
  bind_cols(
    branch_values,
    predict(mult_fit, new_data = branch_values, type = "prob"),
    predict(mult_fit, new_data = branch_values, type = "conf_int", std_error = TRUE)
  ) %>% 
  clean_names() %>% 
  #rename(
  #  target_level = `.Pred_Target level`,
  #  not_target_level = `.Pred_Not target level`,
  #) %>% 
  #left_join(n_df, by = "ee_class") %>% 
  #left_join(ee_class_map %>% distinct(ee_class, ee_order), by = "ee_class") %>%
  mutate(
    branch = branch %>% fct_reorder(pred_positive),
  )
  
plot <- 
  ggplot(
    data = branch_pred, 
    aes(x = pred_positive, y = branch)
  ) +
  geom_errorbar(
    aes(
      xmin = pred_lower_positive,
      xmax = pred_upper_positive,
      colour = geo
    ),
    alpha = 0.4, size = 1, width = 0.1
  ) +
  geom_line(aes(color = geo)) +
  geom_point(aes(color = geo)) +
  labs(
    title = log_df$q_short_e[log_df$question == q] %>% str_wrap(50),
    y = NULL, x = NULL,
    colour = NULL, fill = NULL
  ) +
  #coord_flip() +
  scale_color_viridis_d(end = 0.6, aesthetics = c("colour", "fill")) +
  scale_x_continuous(limits = c(0,1), labels = percent_format()) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    axis.ticks = element_line()
  )

return(plot)

}

plot_q17 <- model_branch(q = "Q17")
plot_q18e <- model_branch(q = "Q18e")
plot_q70c <- model_branch(q = "Q70c")
plot_q70f <- model_branch(q = "Q70f")


```

```{r, fig.width=14, fig.height=10}

plot_row <- plot_grid(plot_q17, plot_q18e, plot_q70c, plot_q70f, nrow = 2)

title <- ggdraw() + 
  draw_label(
    paste0("Effect of Branch on Perception of Workload"),
    fontface = 'bold',
    size = 18,
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )

subtitle <- ggdraw() + 
  draw_label(
    "Predicted probability of positive responses based on Bayesian logistical regression on branch and geographic status. The errorbars display the 90% credible interval.",
    size = 12,
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )

plot <- plot_grid(
  title, subtitle, plot_row,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 0.05, 1)
)

ggsave2(
  plot = plot,
  file.path("plots", paste0("PSES2020@GAC - Effect of Branch on Perception of Workload.png")), 
  width = 14, height = 10
)


```


